\bf Adaption of the Chumbley Score to matching of bullet striation marks

### 2018-03-22 doc/paper.Rmd
couple of edits

\hh{More on NIST database, some on CSAFE}

@@

Lands for all Hamby-44 and Hamby-252 scans are made available through the NIST ballistics database \citep{nist} and were considered, here. Both of these sets of scans are part of the larger Hamby study \citep{hamby} and each consist of twenty known bullets (two each from ten consecutively rifled Ruger P85 barrels) and fifteen questioned bullets (each matching one of the ten barrels). Ground truth for both of these Hamby sets is known and was used to assess correctness of the tests results.

Profiles and signatures were extracted from the Hamby 44 and Hamby 252 data as described in \citet{aoas}.

@@

\hh{For signatures from NIST scans we see three problems: 
\begin{enumerate}
\item type-2 error rate is at best 30\% for a type-1 error rate of 5\%, which is well above the error rates we see for tool marks from screw drivers, see figure \ref{fig:type2};
\item the observed type-1 error, which generally close to the nominal type-1 error rate, depends on the size of the optimization window: as the window size increases, the observed type-1 error decreases, see figure \ref{fig:type1};
\item the Chumbley-score fails to provide a result for up to 3\% of the cases. This drop-out rate is considerably higher when the two lands are from same source than when the lands are from different sources, see figure \ref{fig:missings}. 
\end{enumerate}
}

Following on similar lines to the setup of toolmarks, the first step here is to first identify what difference does different window sizes of optimization and the validation step have, when adapting the toolmark method to bullets.

@@

### Signatures

@@

```
Signatures of lands for all Hamby-44 and Hamby-252 scans made available through the NIST ballistics database \citep{nist} were considered. Both of these sets of scans are part of the larger Hamby study \citep{hamby} and each consist of twenty known bullets (two each from ten consecutively rifled Ruger P85 barrels) and fifteen questioned bullets (each matching one of the ten barrels). Ground truth for both of these Hamby sets is known and was used to assess correctness of the tests results.

@@

The idea therefore is to first use these raw values of the profile directly in the chumbley algorithm, and see how the algorithm performs for different coarseness values (smoothing parameter as referred in the function lowess used in the chumbley algorithm).

\pagebreak

# Results
We used the adjusted Chumbley method as proposed in \citet{hadler} and implemented in the R package `toolmaRk` \citep{toolmark} on all pairwise land-to-land comparisons of the Hamby scans (a total of 85,491 comparisons) with the pairwise sets for the comparisons given in the table \ref{tab:param}.
```{r param}
param_setting<- all %>% select(wo, wv) %>% group_by(wo, wv) %>% tally() %>% select(wo,wv)
#param_setting<- data.frame(matrix(param_setting, nrow = 16))
#t(param_setting)
#param_setting %>% tally() 
library(kableExtra)
#knitr::kable(t(param_setting)[,1:17], digits=0, caption = "Overview of parameter settings used for optimization and validation windows for bullet land signatures.", format = "latex" ,booktabs=TRUE) %>% kableExtra::kable_styling(latex_options = c("scale_down", "hold"))

@@

## Signatures

Figure \ref{fig:type2} gives an overview of type II error rates observed when varying the window size in the optimization step. Two levels of validation window size 30 and 50 were chosen as to compare the error rates for different nominal type I errors. We notice that the trends for these nominal type I errors are similar and in most cases a validation window of 50 has higher type II error than for 30. A change in this trend is seen for a 0.05 $\alpha$ level, although the difference between the two windows is very small for this case. We can also notice an obvious trend of increase in the Type II error as the window of optimization increases and see a minimum around the optimization window size of 120 pixels. Hence we are inclined to choose a smaller validation window size and optimization window as 120.

@@

```{r lms, results='asis'}
nsnest <- ns %>% filter(wv %in% c(30,50)) %>% group_by(match, wv) %>% nest()
nsnest <- nsnest %>% mutate(
  model = data %>% purrr::map(.f = function(d) lm(I(missperc*100)~wo, data=d)),
  coefs = model %>% purrr::map(.f = broom::tidy)
)
tab <- nsnest %>% unnest(coefs) %>% filter(term=="wo") %>% arrange(match) %>%
  select(-term, -statistic, -p.value)
knitr::kable(tab, digits=3, caption = "Estimates of the increase in percent of failed tests corresponding to a 100 point increase in the optimization window.", booktabs=TRUE)
```
Figure \ref{fig:missings} gives an overview of the number of failed tests, i.e. tests in which a particular parameter setting did not return a valid result. This happens, when the shift to align two signatures is so large, that the remaining overlap is too small to accommodate  windows for validation. The problem is therefore exacerbated by a larger validation window. Figure \ref{fig:missings} also shows that the number of failed tests is approximately linear in the size of the optimization window. Test results from different sources have a much higher chance to fail, raising the question, whether failed tests should be treated as rejections of the null hypothesis of same source. For known non-matches there is a higher possibility that in the optimization pair of windows where cross-correlations are maximum are too far apart, and same shifts of this order hit the end of the signature.

>>>>

@@

Lands for all Hamby-44 and Hamby-252 scans are made available through the NIST ballistics database \citep{nist} and are considered, here. Both of these sets of scans are part of the larger Hamby study \citep{hamby}. Each set consists of twenty known bullets (two each from ten consecutively rifled Ruger P85 barrels) and fifteen questioned bullets (each matching one of the ten barrels). Ground truth for both of these Hamby sets is known and was used to assess correctness of the tests results.

\hh{Discuss CSAFE scans}

Profiles and signatures were extracted from all scans as described in \citet{aoas}.

@@

\hh{For signatures from NIST scans we see three problems: 
\begin{enumerate}
\item type-2 error rate is at best 30\% for a type-1 error rate of 5\%, which is well above the error rates we see for tool marks from screw drivers, see figure \ref{fig:type2};
\item the observed type-1 error, which generally close to the nominal type-1 error rate, depends on the size of the optimization window: as the window size increases, the observed type-1 error decreases, see figure \ref{fig:type1};
\item the Chumbley-score fails to provide a result for up to 3\% of the cases. The number of failed tests increases linearly in the size used for the window in the optimization step. The rate of failed tests is considerably higher when the two lands are from same source than when the lands are from different sources, see figure \ref{fig:missings}. 
\end{enumerate}
}

Figure \ref{fig:missings} gives an overview of the number of failed tests, i.e. tests in which a particular parameter setting did not return a valid result. This happens e.g.\ when the shift to align two markings is so large, that the remaining overlap is too small to accommodate  windows for validation. The problem is therefore exacerbated by a larger validation window. Figure \ref{fig:missings}(left) also shows that the number of failed tests is approximately linear in the size of the optimization window.
Tests also fail at a higher rate than expected when the markings are from the same source (right). This difference is the least pronounced around an optimized window size $w_o$ of around 120. However, even in this scenario, the number of failed tests for markings from the same source is about twice as high as expected given the number of same source and different source pairings in the data set.

<!--
Following on similar lines to the setup of toolmarks, the first step here is to first identify what difference does different window sizes of optimization and the validation step have, when adapting the toolmark method to bullets.

@@

### Signatures
-->

@@

<!--
Signatures of lands for all Hamby-44 and Hamby-252 scans made available through the NIST ballistics database \citep{nist} were considered. Both of these sets of scans are part of the larger Hamby study \citep{hamby} and each consist of twenty known bullets (two each from ten consecutively rifled Ruger P85 barrels) and fifteen questioned bullets (each matching one of the ten barrels). Ground truth for both of these Hamby sets is known and was used to assess correctness of the tests results.

@@

The idea therefore is to first use these raw values of the profile directly in the chumbley algorithm, and see how the algorithm performs for different coarseness values (smoothing parameter as referred in the function lowess used in the chumbley algorithm).
-->

# Results
<!--
We used the adjusted Chumbley method as proposed in \citet{hadler} and implemented in the R package `toolmaRk` \citep{toolmark} on all pairwise land-to-land comparisons of the Hamby scans (a total of 85,491 comparisons) with the pairwise sets for the comparisons given in the table \ref{tab:param}.
```{r param}
param_setting<- all %>% select(wo, wv) %>% group_by(wo, wv) %>% tally() %>% select(wo,wv)
#param_setting<- data.frame(matrix(param_setting, nrow = 16))
#t(param_setting)
#param_setting %>% tally() 
library(kableExtra)
#knitr::kable(t(param_setting)[,1:17], digits=0, caption = "Overview of parameter settings used for optimization and validation windows for bullet land signatures.", format = "latex" ,booktabs=TRUE) %>% kableExtra::kable_styling(latex_options = c("scale_down", "hold"))

@@

## Signatures
-->
<!--
Figure \ref{fig:type2} gives an overview of type II error rates observed when varying the window size in the optimization step. Two levels of validation window size 30 and 50 were chosen as to compare the error rates for different nominal type I errors. We notice that the trends for these nominal type I errors are similar and in most cases a validation window of 50 has higher type II error than for 30. A change in this trend is seen for a 0.05 $\alpha$ level, although the difference between the two windows is very small for this case. We can also notice an obvious trend of increase in the Type II error as the window of optimization increases and see a minimum around the optimization window size of 120 pixels. Hence we are inclined to choose a smaller validation window size and optimization window as 120.

@@

```{r lms, results='asis'}
nsnest <- ns %>% filter(wv %in% c(30,50)) %>% group_by(match, wv) %>% nest()
nsnest <- nsnest %>% mutate(
  model = data %>% purrr::map(.f = function(d) lm(I(missperc*100)~wo, data=d)),
  coefs = model %>% purrr::map(.f = broom::tidy)
)
tab <- nsnest %>% unnest(coefs) %>% filter(term=="wo") %>% arrange(match) %>%
  select(-term, -statistic, -p.value)
knitr::kable(tab, digits=3, caption = "Estimates of the increase in percent of failed tests corresponding to a 100 point increase in the optimization window.", booktabs=TRUE)
```
-->
<!--
Figure \ref{fig:missings} gives an overview of the number of failed tests, i.e. tests in which a particular parameter setting did not return a valid result. This happens, when the shift to align two signatures is so large, that the remaining overlap is too small to accommodate  windows for validation. The problem is therefore exacerbated by a larger validation window. Figure \ref{fig:missings} also shows that the number of failed tests is approximately linear in the size of the optimization window. Test results from different sources have a much higher chance to fail, raising the question, whether failed tests should be treated as rejections of the null hypothesis of same source. For known non-matches there is a higher possibility that in the optimization pair of windows where cross-correlations are maximum are too far apart, and same shifts of this order hit the end of the signature. 
-->

### 2018-03-22 doc/paper.tex
couple of edits

\hh{For signatures from NIST scans we see three problems: 
\begin{enumerate}
\item type-2 error rate is at best 30\% for a type-1 error rate of 5\%, which is well above the error rates we see for tool marks from screw drivers, see figure \ref{fig:type2};
\item the observed type-1 error, which generally close to the nominal type-1 error rate, depends on the size of the optimization window: as the window size increases, the observed type-1 error decreases, see figure \ref{fig:type1};
\item the Chumbley-score fails to provide a result for up to 3\% of the cases. This drop-out rate is considerably higher when the two lands are from same source than when the lands are from different sources, see figure \ref{fig:missings}. 
\end{enumerate}
}

Following on similar lines to the setup of toolmarks, the first step
here is to first identify what difference does different window sizes of
optimization and the validation step have, when adapting the toolmark
method to bullets.

The marking made on bullets are smaller than toolmarks and is also less
wider. The idea is to find out possible areas of error while adapting
the score based method proposed for toolmarks, using cross-validation
setup to identify appropriate parameter settings for (a) signatures and
(b) profiles directly

\subsubsection{Signatures}\label{signatures}

@@

Signatures of lands for all Hamby-44 and Hamby-252 scans made available
through the NIST ballistics database \citep{nist} were considered. Both
of these sets of scans are part of the larger Hamby study \citep{hamby}
and each consist of twenty known bullets (two each from ten
consecutively rifled Ruger P85 barrels) and fifteen questioned bullets
(each matching one of the ten barrels). Ground truth for both of these
Hamby sets is known and was used to assess correctness of the tests
results.

Bullet signatures being compared at this time are therefore from the
Hamby 44 and Hamby 252 data. The database setup and pre-processing
system used for choosing the Bullet signatures are as described by
\citet{aoas}. In order to choose the bullet signatures we first filter
out Land\_id for Profiles from the Hamby 44 and Hamby 252 data and
remove all NA values. Then run\_id = 3 is chosen as the signatures
generated from this run\_id give the closest match. Different run\_id's
have some different settings for generating the signatures.(The level of
smoothing does not seem to be one of them)

The bullet signatures when generated by this process already includes a
loess smoothing. Therefore, the coarseness factor is set to 1 while
running the chumbley non random algorithm for comparing different
optimization windows.The algorithm generates the same\_shift,
different\_shift, U-Stat and P\_value parameters which are then used to
calculate the errors associated with different sets of window sizes.

\subsubsection{Profiles}\label{profiles}

The profiles are cross-sectional values of the the bullet striation mark
which are chosen at an optimum height (x as used by \citet{aoas}). This
x or height is not a randomly chosen level. The rationale behind the
choice has been explained by \citet{aoas}. A region is first chosen
where the cross-correlation seems to change very less and in this region
an optimum height is chosen. The profiles generally resemble a curve
which is more or less similar to a quadratic curve (a quadratic fit to
the raw data values of the profile is not an exact fit but it does show
a similar trend). Profiles are the set of raw values representing the
striation marks, and signatures are generated from these by removal of
the inherent curvature and applying some smoothing (the signatures
generated by \citet{aoas} use a loess function for smoothing).

Similar to signatures the run\_id = 3 was used when applying the
chumbley algorithm using the database setup given by \citet{aoas} of
Hamby-44 and Hamby-252 datasets, on the profiles. The run\_id not only
defines the level of smoothing but also signifies the chosen height at
which the profiles were selected initially. Another important aspect is
the range of horizontal values (which is referred to as the y values in
\citet{aoas}) in the signatures. These have already been pre-processed
in the database to not include any grooves.

Therefore for the sake of comparison the run\_id = 3 is still chosen so
as to ensure that the horizontal values remain the same as that of the
signatures. This also gives us profiles with the grooves removed.

The idea therefore is to first use these raw values of the profile
directly in the chumbley algorithm, and see how the algorithm performs
for different coarseness values (smoothing parameter as referred in the
function lowess used in the chumbley algorithm).

@@

We used the adjusted Chumbley method as proposed in \citet{hadler} and
implemented in the R package \texttt{toolmaRk} \citep{toolmark} on all
pairwise land-to-land comparisons of the Hamby scans (a total of 85,491
comparisons) with the pairwise sets for the comparisons given in the
table \ref{tab:param}.

\begin{table}[!h][!h]
\caption{\label{tab:param}Overview of parameter settings used for optimization and validation windows for bullet land signatures.}

\centering
\resizebox{\linewidth}{!}{\resizebox{\linewidth}{!}{\begin{tabular}[t]{lrrrrrrrrrrrrrrrrrr}
\toprule
wo & 50 & 50 & 60 & 60 & 80 & 80 & 90 & 90 & 100 & 100 & 110 & 110 & 120 & 120 & 120 & 120 & 120 & 120\\
wv & 30 & 50 & 30 & 50 & 30 & 50 & 30 & 50 & 30 & 50 & 30 & 50 & 10 & 20 & 30 & 40 & 50 & 60\\
\bottomrule
\end{tabular}}}
\centering
\begin{tabular}[t]{lrrrrrrrrrrrrrrr}
\toprule
wo & 130 & 130 & 140 & 140 & 150 & 150 & 160 & 160 & 200 & 200 & 200 & 240 & 240 & 280 & 280\\
wv & 30 & 50 & 30 & 50 & 30 & 50 & 30 & 50 & 20 & 30 & 50 & 30 & 50 & 30 & 50\\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[!h]
\caption{\label{tab:confusion}Confusion Table for different optimization window sizes with validation window size as 30.}

\centering
\begin{tabular}[t]{llrl}
\toprule
\hspace{1em}\hspace{1em}signif & match & Freq & Type\\
\midrule
\addlinespace[0.5em]
\multicolumn{4}{l}{\textbf{Size of Optimization Window = 280}}\\
\hspace{1em}FALSE & FALSE & 78909 & True Negative\\
\hspace{1em}TRUE & FALSE & 4192 & False Positive (Type I)\\
\hspace{1em}FALSE & TRUE & 446 & False Negative (Type II)\\
\hspace{1em}TRUE & TRUE & 731 & True Positive\\
\bottomrule
\end{tabular}
\centering
\begin{tabular}[t]{llrl}
\toprule
signif & match & Freq & Type\\
\midrule
\addlinespace[0.5em]
\multicolumn{4}{l}{\textbf{Size of Optimization Window = 120}}\\
\hspace{1em}FALSE & FALSE & 79249 & True Negative\\
\hspace{1em}TRUE & FALSE & 4523 & False Positive (Type I)\\
\hspace{1em}FALSE & TRUE & 386 & False Negative (Type II)\\
\hspace{1em}TRUE & TRUE & 854 & True Positive\\
\bottomrule
\end{tabular}
\centering
\begin{tabular}[t]{llrl}
\toprule
signif & match & Freq & Type\\
\midrule
\addlinespace[0.5em]
\multicolumn{4}{l}{\textbf{Size of Optimization Window = 80}}\\
\hspace{1em}FALSE & FALSE & 79477 & True Negative\\
\hspace{1em}TRUE & FALSE & 4503 & False Positive (Type I)\\
\hspace{1em}FALSE & TRUE & 463 & False Negative (Type II)\\
\hspace{1em}TRUE & TRUE & 781 & True Positive\\
\bottomrule
\end{tabular}
\end{table}

\subsection{Signatures}\label{signatures-1}

Figure \ref{fig:type2} gives an overview of type II error rates observed
when varying the window size in the optimization step. Two levels of
validation window size 30 and 50 were chosen as to compare the error
rates for different nominal type I errors. We notice that the trends for
these nominal type I errors are similar and in most cases a validation
window of 50 has higher type II error than for 30. A change in this
trend is seen for a 0.05 \(\alpha\) level, although the difference
between the two windows is very small for this case. We can also notice
an obvious trend of increase in the Type II error as the window of
optimization increases and see a minimum around the optimization window
size of 120 pixels. Hence we are inclined to choose a smaller validation
window size and optimization window as 120.

Table \ref{tab:confusion} shows the confusion tables with the
classification of type I and type II errors and how the numbers change
with a change in the optimization window. The windows represent areas to
the left of the window with minimum type II, near the minimum type II
window and to the far right of the minimum type II error

Figure \ref{fig:type1} compares nominal (fixed) type I error and
actually observed type I errors for the parameter settings in table
\ref{tab:param}. With an increasing size of the window used in the
optimization step the observed type I error rate decreases (slighty).
This means as the optimization window increase the observed type I error
rate gets smaller. A smaller validation window on the other hand, tends
to be associated with a higher type I error rate. This can be better
imagined for a given window of optimization, where the actual Type I
error is comparable to the nominal level for only a select few
validation window sizes. For these comparable validation window sizes of
30 and 50 as done here, the actual type I error increases very slightly
and can be seen in Figure \ref{fig:type1}. This increase is not as much
when compared to the variation seen with the optimization window sizes.
This effect might be related to the increasing number of tests that fail
for larger optimization window sizes, in particular for non-matching
striae (see fig \ref{fig:missings}).

{\centering \includegraphics[width=\textwidth]{figures/wo120-1}

\caption{The figure on the left shows the actual Type I error while the figure on the right shows the Type II error for different validation window sizes and different chosen nominal alpha levels when the size of the optimization window = 120}\label{fig:wo120}
\end{figure}

The actual type I error and type II error for signatures were also
compared for different validation window sizes. Figure \ref{fig:wo120}
shows the actual rates for different nominal \(\alpha\) levels. We can
see that the type II error rises with higher validation windows for the
smaller nominal \(\alpha\) levels while for the nominal \(\alpha\) =
0.05 its almost constant.

\begin{table}

\caption{\label{tab:lms}Estimates of the increase in percent of failed tests corresponding to a 100 point increase in the optimization window.}
\centering
\begin{tabular}[t]{lrrr}
\toprule
match & wv & estimate & std.error\\
\midrule
FALSE & 30 & 0.510 & 0.029\\
FALSE & 50 & 0.936 & 0.036\\
TRUE & 30 & 2.933 & 0.247\\
TRUE & 50 & 3.843 & 0.321\\
\bottomrule
\end{tabular}
\end{table}

Figure \ref{fig:missings} gives an overview of the number of failed
tests, i.e.~tests in which a particular parameter setting did not return
a valid result. This happens, when the shift to align two signatures is
so large, that the remaining overlap is too small to accommodate windows
for validation. The problem is therefore exacerbated by a larger
validation window. Figure \ref{fig:missings} also shows that the number
of failed tests is approximately linear in the size of the optimization
window. Test results from different sources have a much higher chance to
fail, raising the question, whether failed tests should be treated as
rejections of the null hypothesis of same source. For known non-matches
there is a higher possibility that in the optimization pair of windows
where cross-correlations are maximum are too far apart, and same shifts
of this order hit the end of the signature.

\subsection{Profiles}\label{profiles-1}

>>>>

\hh{For signatures from NIST scans we see three problems: 
\begin{enumerate}
\item type-2 error rate is at best 30\% for a type-1 error rate of 5\%, which is well above the error rates we see for tool marks from screw drivers, see figure \ref{fig:type2};
\item the observed type-1 error, which generally close to the nominal type-1 error rate, depends on the size of the optimization window: as the window size increases, the observed type-1 error decreases, see figure \ref{fig:type1};
\item the Chumbley-score fails to provide a result for up to 3\% of the cases. The number of failed tests increases linearly in the size used for the window in the optimization step. The rate of failed tests is considerably higher when the two lands are from same source than when the lands are from different sources, see figure \ref{fig:missings}. 
\end{enumerate}
}

Figure \ref{fig:missings} gives an overview of the number of failed
tests, i.e.~tests in which a particular parameter setting did not return
a valid result. This happens e.g.~when the shift to align two markings
is so large, that the remaining overlap is too small to accommodate
windows for validation. The problem is therefore exacerbated by a larger
validation window. Figure \ref{fig:missings}(left) also shows that the
number of failed tests is approximately linear in the size of the
optimization window. Tests also fail at a higher rate than expected when
the markings are from the same source (right). This difference is the
least pronounced around an optimized window size \(w_o\) of around 120.
However, even in this scenario, the number of failed tests for markings
from the same source is about twice as high as expected given the number
of same source and different source pairings in the data set.

@@

@@

\subsection{Profiles}\label{profiles}
